---
title: "Lab07"
author: "Doan Cong Thanh - 22125096"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE, message = FALSE)
```
## Question 5

### Set up
External library
```{r}
library(car)
```

Data set
```{r}
data5 <- read.csv("manuf.csv", header = TRUE)
```
```{r echo=FALSE}
str(data5)
```   

Model: $Y=\alpha K^{\beta_2} L^{\beta_3} E^{\beta_4} M^{\beta_5} exp\{\}$
$\\ \Rightarrow ln{Y} = ln{A}+\beta_2ln{K}+\beta_3ln{L}+\beta_4ln{E}+\beta_5ln{M}+e$

```{r}
log_y <- log(data5$y)
log_k <- log(data5$k)
log_l <- log(data5$l)
log_e <- log(data5$e)
log_m <- log(data5$m)
model5 <- lm(log_y ~ log_k + log_l + log_e + log_m)
```

Summary
```{r,echo=FALSE}
summary(model5)
```

### a)
$$
\begin{cases}
H_0:\beta_2=0
\\H_1:\beta_2 \neq 0
\end{cases}
$$
Apply Student's test:  
Assume $\alpha=0.05$  
From the table of summary, we obtain $p-value = 0.8310$  
Since $p-value > \alpha$ $(0.8310 > 0.05)$  
$\Rightarrow$ Fail to reject $H_0$

### b)
$$
\begin{cases}
H_0:\beta_2=\beta_3=0
\\H_1: \exists i|\beta_i \neq 0
\end{cases}
$$
When $H_0$ is true, we have reduced model:
$ln{Y} = ln{A}+\beta_4ln{E}+\beta_5ln{M}+e$

```{r}
model_reduce5b <- lm(log_y ~ log_e + log_m)
anova(model_reduce5b, model5)
```

Apply the Partial Fisher test:\
Assume $\alpha=0.05$\
From the ANNOVA table, we obtain: $p-value=0.8616417$\
Since $p-value>\alpha$ $(0.8616417>0.05)$
$\Rightarrow$ Fail to reject $H_0$

### c)
$$
\begin{cases}
H_0:\beta_2=\beta_4=0
\\H_1: \exists i|\beta_i \neq 0
\end{cases}
$$
When $H_0$ is true, we have reduced model:
$ln{Y} = ln{A}+\beta_3ln{L}+\beta_5ln{M}+e$

```{r}
model_reduce5c <- lm(log_y ~ log_l + log_m)
anova(model_reduce5c, model5)
```

Apply the Partial Fisher test:\
Assume $\alpha=0.05$\
From the ANNOVA table, we obtain: $p-value=0.8810752$\
Since $p-value>\alpha$ $(0.8810752>0.05)$
$\Rightarrow$ Fail to reject $H_0$

### d)
$$
\begin{cases}
H_0:\beta_2=\beta_3=\beta_4=0
\\H_1: \exists i|\beta_i \neq 0
\end{cases}
$$
When $H_0$ is true, we have reduced model:
$ln{Y} = ln{A}+\beta_5ln{M}+e$

```{r}
model_reduce5d <- lm(log_y ~ log_m)
anova(model_reduce5d, model5)
```

Apply the Partial Fisher test:\
Assume $\alpha=0.05$\
From the ANNOVA table, we obtain: $p-value=0.9077512$\
Since $p-value>\alpha$ $(0.9077512>0.05)$
$\Rightarrow$ Fail to reject $H_0$

### e)
$$
\begin{cases}
H_0:\beta_2+\beta_3+\beta_4+\beta_5=1
\\H_1: \beta_2+\beta_3+\beta_4+\beta_5 \neq  1
\end{cases}
$$

```{r}
coef <- coef(model5)
vcov <- vcov(model5)
R <- c(0, 1, 1, 1, 1)
k <- 1
Rb <- sum(coef[2:5])
SE <- sqrt(vcov[2, 2] + vcov[3, 3] + vcov[4, 4] + vcov[5, 5]
 + 2 * (vcov[2, 3] + vcov[2, 4] + vcov[2, 5]
 + vcov[3, 4] + vcov[3, 5] + vcov[4, 5]))
t_0 <- (sum(coef[2:5]) - k) / SE
t_crit <- qt(0.975, 20)
```

```{r}
t_0
```

```{r}
t_crit
```

Apply Student's test:\
We obtain $t_{crit} = 2.085963, t_0=-0.0254$\
Since $|t_0|<t_{crit}$\
$\Rightarrow$Reject $H_0$

### f)

```{r}
vif(model5)
```
 From the VIF table, we observe significant high values of all the variables, which indicates severe multicollinearity in this model.\
 High multicollinearity makes our estimate unpredictable and unstable, moreover, the model becomes highly sensitive with a small change in data. Therefore, we need to eliminate some variables, starting from the largest one, then calculate and reconsider again the VIF table.
 
```{r}
model_5f1 <- lm(log_y ~ log_k + log_l + log_m)
vif(model_5f1)
```

```{r}
model_5f2 <- lm(log_y ~ log_k + log_l)
vif(model_5f2)
```
In this case, we compare two simple model (model with only 1 variable).

```{r}
model_5f3 <- lm(log_y ~ log_k)
summary(model_5f3)
```

```{r}
model_5f4 <- lm(log_y ~ log_l)
summary(model_5f4)
```   

Since model_5f4 hold the large values for Multiple $R^2$ and Adjusted $R^2$. So this is the best model.

## Question 6
### a)
```{r}
data6 <- read.csv("stockton4.csv", header = TRUE)
str(data6)
attach(data6)
```

$$
SPRICE = \beta_1 + \beta_2 LIVAREA + \beta_3 AGE + \beta_3 BEDS + \beta_4 BATHS + e
$$
```{r}
model6 <- lm(sprice ~ livarea + age + beds + baths)
summary(model6)
```

Interpret coefficient:

$\beta_1$: In an ideal condition, when all independent varibles equal 0. The the selling price of home will remain 11154.29 dollars.\
$\beta_2$: The selling price of home will increase 10680 dollars corresponding to an increasing of 1 hundred of square feet of the living area.\
$\beta_3$: The selling price of home will decrease 11.33 corresponding to an increasing of 1 year of the home age at time of sale.\
$\beta_4$: The selling price of home will decrease 15552.44 corresponding to an increasing of 1 bed of the home.\
$\beta_5$: The selling price of home will decrease 7019.3 corresponding to an increasing of 1 bath of the home.

### b)
$SPRICE(AGE=10) - SPRICE(AGE=2) = \beta_3 \times 8$

First, we need to calculate the $95\%$ CI for $\beta_3$
```{r}
confint(model6, 3)
```
Hence $95\%$ interval estimate of the difference (in dollar) is:
$8 \times [-169.2429,146.575] =  [-1353.94, 1172.6]$

### c)
Expected increase in price she will get from this extension:
$SPRICE(LIVAREA = 22) - SPRICE(LIVAREA = 20) = \hat\beta_3 \times 2 = 10680 \times 2 = 21360$ dollars

$$
\begin{cases}
H_0:2 \times \beta_2=20000 \Rightarrow \beta_2=10000
\\H_1: \beta_2<10000
\end{cases}
$$
Apply Student's test:
Calculate T-statistic:
$$
T_0 = \frac{\hat{\beta_2}-10000} {S_{\hat{\beta_2}}} = 2.4895
$$
Assume $\alpha = 0.05$, with $n=1500$, which is quite large, so we can use $z-statistic$ instead.
We have $t_{crit} = t_{0.05,1495} = z_0.05 = 1.645$

We reject $H_0$ if $T_0 < -t_{crit}$\
$\Rightarrow$ Fail to reject $H_0$\
Hence, there is enough evidence that the price will be at least $20000.

### d)
Expected increase in price she will get from this extension:\
$2 \times \hat\beta_2 + \hat\beta_4 = 5807.56$\
Expected price increase: $2 \times \beta_2 + \beta_4$\
```{r}
df = 1495
t_crit = qt(1-0.05/2, df)
se = sqrt(4*vcov(model6)["livarea", "livarea"] + vcov(model6)["beds", "beds"] + 2*2*vcov(model6)["livarea","beds"])

lower_bound = 5807.56 - t_crit * se
upper_bound = 5807.56 + t_crit * se
```

```{r}
lower_bound
```

```{r}
upper_bound
```

$95\%$ interval estimate for the expected price: $[2139.593, 9475.527]$

## Question 7
### a)
```{r}
data7 <- read.csv("stockton4.csv", header = TRUE)
str(data7)
detach(data6)
attach(data7)
```

Model:
$ln(SPRICE) = \beta_1 + \beta_2 LIVAREA + \beta_3 LIVAREA^2 + \beta_4 AGE + \beta_5 AGE^2 + \beta_6 BEDS + \beta_7(LIVAREA \times BEDS) + \beta_8(LIVAREA^2 \times BEDS) + \beta_9(AGE \times BEDS) + \beta_{10}(AGE^2 \times BEDS) + e$
```{r}
model7 <- lm(I(log(sprice)) ~ livarea + I(livarea^2) + age + I(age^2) + beds 
 + I(livarea*beds) + I(livarea^2 * beds) + I(age * beds) + I(age^2 * beds))
summary(model7)
```

Estimated relationship between $ln(SPRICE), LIVAREA, and AGE$:\
For 2 bedrooms:\
$ln(SPRICE) = \beta_1 + \beta_2 LIVAREA + \beta_3 LIVAREA^2 + \beta_4 AGE + \beta_5 AGE^2 + 2 \beta_6 + 2 \beta_7 LIVAREA + 2 \beta_8 LIVAREA^2 + 2 \beta_9 AGE^2 + 2 \beta_{10} AGE^2$\
For 3 bedrooms:\
$ln(SPRICE) = \beta_1 + \beta_2 LIVAREA + \beta_3 LIVAREA^2 + \beta_4 AGE + \beta_5 AGE^2 + 3 \beta_6 + 3 \beta_7 LIVAREA + 3 \beta_8 LIVAREA^2 + 3 \beta_9 AGE^2 + 3 \beta_{10} AGE^2$\
For 4 bedrooms:\
$ln(SPRICE) = \beta_1 + \beta_2 LIVAREA + \beta_3 LIVAREA^2 + \beta_4 AGE + \beta_5 AGE^2 + 4 \times \beta_6 + 4 \beta_7 LIVAREA + 4 \beta_8 LIVAREA^2 + 4 \beta_9 AGE^2 + 4 \beta_{10} AGE^2$\

### b)
$$
\begin{cases}
H_0: \beta_6 = \beta_8 = \beta_9 = \beta_{10} = 0\\
H_1: \exists i | \beta_i \neq 0
\end{cases}
$$
When $H_0$ is true, we have reduced model:\
$ln(SPRICE) = \beta_1 + \beta_2 LIVAREA + \beta_3 LIVAREA^2 + \beta_4 AGE + \beta_5 AGE^2 + \beta_7(LIVAREA \times BEDS) + e$
```{r}
reduced_model7 <- lm(I(log(sprice)) ~ livarea + I(livarea^2) + age + I(age^2) + I(livarea*beds))
anova(reduced_model7, model7)
```

Apply the Partial Fisher-testing:
Assume $\alpha = 0.05$\
From the table, we obtain $p-value = 0.32$
Since $p-value > \alpha$\
$\Rightarrow$ Fail to reject $H_0$

### c)
```{r}
model7c <- reduced_model7
summary(model7c)
```

Estimated relationship between ln(sprice), livarea, and age:\
For 2 bedrooms:\
$ln(SPRICE) = \beta_1 + \beta_2 LIVAREA + \beta_3 LIVAREA^2 + \beta_4 AGE + \beta_5 AGE^2 + 2 \beta_7 LIVAREA$\
For 3 bedrooms:\
$ln(SPRICE) = \beta_1 + \beta_2 LIVAREA + \beta_3 LIVAREA^2 + \beta_4 AGE + \beta_5 AGE^2 + 3 \beta_7 LIVAREA$\
For 4 bedrooms:\
$ln(SPRICE) = \beta_1 + \beta_2 LIVAREA + \beta_3 LIVAREA^2 + \beta_4 AGE + \beta_5 AGE^2 + 4 \beta_7 LIVAREA$\


### d)
```{r}
fullModel <- model7
reducedModel <- model7c
```

#### 1) AIC
```{r}
AIC(fullModel)
```

```{r}
AIC(reducedModel)
```

A lower AIC value indicates a model that has a better balance between goodness of fit and complexity.
Since the reduced model has a lower AIC (-342.5015) compared to the full model (-339.2205), the reduced model is chosen.

#### 2) BIC
```{r}
BIC(fullModel)
```

```{r}
BIC(reducedModel)
```

Since the reduced model has a lower BIC (-305.309) compared to the full model (-280.7751), the reduced model is preferred according to the BIC criterion as well.
