\documentclass[10pt,twocolumn]{article}
\usepackage[margin=2cm]{geometry}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{float}
\usepackage{tabularx}
\usepackage{array}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{caption}
\usepackage{xcolor}

\graphicspath{{figures/}}

% ── Space optimisation ──────────────────────────────────────────────────────
\setlength{\emergencystretch}{3em}
\setlength{\parskip}{0.3em}
\setlength{\abovecaptionskip}{4pt}
\setlength{\belowcaptionskip}{2pt}
\setlength{\textfloatsep}{8pt plus 2pt minus 2pt}
\setlength{\floatsep}{6pt plus 2pt minus 2pt}
\setlength{\intextsep}{6pt plus 2pt minus 2pt}
\captionsetup{font=small,labelfont=bf,skip=4pt}
\setlist{nosep,leftmargin=1.2em}

\author{
Loc D. T. Nguyen \and
Nhan T. T. Le \and
Tai Nguyen \\
Faculty of Information Technology \\
VNUHCM -- University of Science, Ho Chi Minh City, Vietnam
}

\title{Evaluating XAI Metrics in Medical AI:\\
A Cross-Study Co-occurrence Analysis and Evaluation Taxonomy}
\date{}

\begin{document}

\maketitle

\begin{abstract}
Explainable Artificial Intelligence (XAI) is increasingly required in clinical settings, yet the metrics used to evaluate explanation quality remain fragmented and rarely compared across medical sub-domains. This paper presents, to our knowledge, the first systematic cross-domain co-occurrence analysis of XAI evaluation metrics in medical AI, examining 52 peer-reviewed studies spanning oncology, cardiology, neuroimaging, radiology, and clinical decision support. We employ a four-phase methodology: systematic corpus collection via PubMed, Scopus, IEEE Xplore, and Google Scholar; criterion extraction and normalisation; cross-domain pairwise correlation analysis; and taxonomy construction. Six quantitative metrics are identified. Faithfulness appears in 47/52 studies (90.4\%), with its absence concentrated in human-centred evaluation designs. Sensitivity co-occurs with Faithfulness in 33 studies (63.5\%). Among perturbation-based metrics, AOPC/MoRF and Pixel Flipping exhibit the strongest correlation ($\rho = 0.51$, $p < 0.001$), indicating partial methodological redundancy. Sufficiency remains isolated (4/52; 7.7\%) despite direct relevance to clinical decision adequacy. Mapping the corpus onto a three-layer taxonomy---mathematical faithfulness, system robustness, and human-centred trust---reveals evaluation effort concentrated in Layer~1 (90.4\%), with robustness (34.6\%) and human-centred criteria (13.5\%) substantially underrepresented. We propose an empirically grounded minimal evaluation framework spanning all three layers and contribute a structured taxonomy validated against cross-domain usage patterns.
\end{abstract}

% ─────────────────────────────────────────────────────────────
\input{intro_back}
\input{related_work}
\input{methodology}
\input{results}
\input{end}

\bibliographystyle{ieeetr}
\bibliography{citation}
% ─────────────────────────────────────────────────────────────

\end{document}
