\section{Results}
% ──────────────────────────────────────────────────────────────────────────────

To evaluate the fragmentation and metric usage patterns within medical XAI,
we applied our four-phase methodology to an expanded corpus of 52 peer-reviewed
dataset papers spanning oncology, cardiology, neurology, radiology, and broader
clinical decision-support settings. The following sections report our findings across corpus
composition, metric co-occurrence, cross-domain alignment, and taxonomy
validation.

% ──────────────────────────────────────
\subsection{Corpus Composition and Metric Distribution}

The corpus spans 52 dataset papers across oncology, radiology, cardiology,
neuroimaging, and broader clinical decision-support settings, with imaging-heavy
applications contributing the largest share.
Table~\ref{tab:dataset_papers} provides a structured
summary of the most representative studies included in this corpus. Across
all papers, six quantitative evaluation metrics were identified:
Faithfulness, Sensitivity, AOPC/MoRF, Pixel Flipping, IROF, and
Sufficiency. Faithfulness was the most frequently reported metric,
appearing in 47 of 52 studies (90\%), followed by Sensitivity in 34 studies
(65\%), Pixel Flipping in 14 studies (27\%), and AOPC/MoRF in 13 studies
(25\%). IROF and Sufficiency each appeared in 4 studies (8\%), the latter
remaining underutilised despite its direct clinical relevance. Crucially,
the five studies in which Faithfulness is absent are concentrated in
human-centred evaluation designs~\cite{rezaeian2026explainability,finzel2024human,bauer2025human}
and consistency-focused benchmarks~\cite{qin2025consistency,alpsalaz2026explainability},
a pattern that reveals a structural separation between technical and
clinician-oriented XAI research rather than a random omission.

% ── Table I: summary of medical XAI literature
\input{tables_I}

% ──────────────────────────────────────
\subsection{Metric Co-occurrence and Correlation}

Examining metric co-occurrence patterns across the reviewed studies
revealed three distinct patterns. First, Faithfulness and Sensitivity
co-occurred in 33 of the 47 studies reporting Faithfulness, confirming a
strong but no longer universal pairing. In the five studies lacking
Faithfulness, Sensitivity still appeared in one case~\cite{qin2025consistency},
indicating that the two metrics can be operationalised independently when
evaluation design prioritises stability over fidelity. Table~\ref{tab:sensitivity_methods}
lists the studies in which sensitivity analysis formed the primary evaluation method.
Second, AOPC/MoRF and Pixel Flipping emerged as the most mechanistically similar
pair, co-occurring in 7 studies and producing the strongest Spearman correlation in the
corpus ($\rho = 0.51$, $p < 0.001$); AOPC and IROF also correlated significantly
($\rho = 0.37$, $p = 0.007$), confirming that perturbation-based metrics cluster together
and likely capture overlapping properties. Third, Sufficiency rarely co-occurred with any
other metric, confirming that it measures a distinct evaluative dimension---namely,
the consistency of explanations across similar patient instances---that the remaining
metrics do not address.

% ── Table II: sensitivity analysis studies

\input{tables_II} 

% ──────────────────────────────────────
\subsection{Cross-Domain Alignment Within Medical Sub-domains}

Laying the normalised criterion list across the five medical sub-domains
revealed both universal and domain-specific patterns. Faithfulness and
Sensitivity qualified as near-universal criteria, appearing in at least
four of the five sub-domains with broadly comparable operationalisations.
Table~\ref{tab:faithfulness_metrics} summarises studies that explicitly
report Faithfulness Correlation or Faithfulness Estimate as their primary
evaluation metric. By contrast, IROF and Pixel Flipping were predominantly
domain-specific, concentrated in radiology and oncology where saliency-map
explanations are standard. The expanded radiology sub-corpus --- now including
high-quality benchmarking studies such as Saporta et al.~\cite{saporta2022benchmarking}
in \textit{Nature Machine Intelligence} and Wollek et al.~\cite{wollek2023attention}
in \textit{Radiology: AI} --- confirms that this domain leads in metric diversity
and methodological rigour. Neuroimaging studies relied predominantly on Sensitivity,
with notably fewer metrics per paper than other sub-domains, indicating that
neurological XAI evaluation remains comparatively underdeveloped; Qin~\cite{qin2025consistency}
represents an emerging exception, applying consistency and stability benchmarking
to brain MRI classification. Oncology studies were the most metrically diverse, combining
technical and, in several cases, user-study-based criteria, with new contributions such
as Cerekci et al.~\cite{cerekci2024quantitative} providing dedicated quantitative evaluation
of saliency methods in mammography. The clinical sub-domain expanded most substantially,
gaining human-centred studies~\cite{rezaeian2026explainability,bauer2025human,finzel2024human}
that operationalise Layer~3 trust criteria previously absent from the corpus.

% ── Table III: faithfulness correlation studies
\input{tables_III} 

% ──────────────────────────────────────
\subsection{Taxonomy Validation Against Medical Corpus}

Mapping the reviewed studies onto the three-layer taxonomy confirmed that
evaluation effort remains heavily concentrated in the first layer.
47 of 52 studies (90\%) reported at least one mathematical faithfulness
criterion, while 18 studies (35\%) addressed system robustness through
multi-metric perturbation analyses, and 7 studies (13\%) included any
human-centred trust measure or sufficiency check. Ihongbe and
Fouad~\cite{IhongbeFouad2024}, Fouad et al.~\cite{11355720},
Rezaeian et al.~\cite{rezaeian2026explainability}, Bauer and
Michalowski~\cite{bauer2025human}, and Finzel~\cite{finzel2024human}
represent the clearest operationalisations of the trust layer, collectively
demonstrating that cognitive load, clinician comprehension, and diagnostic
confidence are measurable but remain minority concerns in the literature.
The distributional imbalance validates the taxonomy's diagnostic utility:
it makes the coverage gap in layers two and three immediately visible.
Table~\ref{tab:perturbation_methods}
lists perturbation-based and pixel-flipping studies, which collectively
represent the most common operationalisation of the faithfulness layer in
medical imaging contexts. Cross-layer criteria, most notably contrastive
explanations, appeared in several papers and were flagged accordingly.
Grad-CAM-based studies consistently reported higher Faithfulness scores than
LIME-based studies across sub-domains, suggesting that the choice of XAI
method itself substantially conditions what evaluation scores are achievable.

% ── Table IV: pixel flipping / perturbation studies

\input{tables_IV}

% ─────────────────────────────────────────────────────────────────────────────
