\documentclass{article}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{float}
\usepackage{longtable}
\usepackage{tabularx}
\usepackage{ltablex}
\usepackage{array}
\usepackage{booktabs}

\setlength{\emergencystretch}{3em}

\author{
Loc D. T. Nguyen \and
Nhan T. T. Le \and
Tai Nguyen \\
Faculty of Information Technology \\
VNUHCM -- University of Science, Ho Chi Minh City, Vietnam
}

\title{Evaluating XAI Metrics in Medical AI:\\
A Cross-Study Co-occurrence Analysis and Evaluation Taxonomy}
\date{}

\begin{document}

\maketitle

% [REVISED ABSTRACT — Q1 STYLE]
\begin{abstract}
Explainable Artificial Intelligence (XAI) is increasingly required in clinical settings to support transparent decision-making. However, the metrics used to evaluate explanation quality in medical AI remain fragmented, inconsistently defined, and rarely compared across clinical sub-domains---creating a significant barrier to reproducible and trustworthy deployment. This paper addresses this gap through a systematic cross-study co-occurrence analysis of XAI evaluation metrics, examining 52 peer-reviewed studies spanning five medical sub-domains: oncology, cardiology, neuroimaging, radiology, and clinical decision support. We employ a four-phase methodology comprising corpus collection via systematic search across PubMed, Scopus, IEEE Xplore, and Google Scholar; criterion extraction and normalisation; cross-domain alignment with pairwise correlation analysis; and taxonomy construction. Six quantitative metrics in active use are identified and analysed. Faithfulness is the most prevalent metric, appearing in 47 of 52 studies (90.4\%), while its absence is concentrated in human-centred evaluation designs---a pattern revealing a structural divide between technical and clinician-oriented XAI research. Sensitivity co-occurs with Faithfulness in 33 studies (63.5\%). Among perturbation-based metrics, AOPC/MoRF and Pixel Flipping exhibit the strongest pairwise correlation ($\rho = 0.51$, $p < 0.001$), suggesting partial redundancy. Sufficiency remains isolated and underutilised (4 of 52 studies; 7.7\%) despite its direct relevance to clinical decision adequacy. Mapping the corpus onto a three-layer taxonomy---mathematical faithfulness, system robustness, and human-centred trust---reveals that evaluation effort is heavily concentrated in the first layer (90.4\%), with robustness addressed in 18 studies (34.6\%) and human-centred criteria in only 7 (13.5\%). Building on these findings, we propose a minimal evaluation framework spanning all three layers, comprising one faithfulness metric, one perturbation-based metric, one sensitivity metric, one sufficiency metric, and one human-centred metric. We further contribute a structured taxonomy grounded in empirical usage patterns rather than purely conceptual design, and identify standardisation priorities for underserved sub-domains such as neuroimaging.
\end{abstract}

% ─────────────────────────────────────────────────────────────
\input{intro_back}
\input{related_work}
\input{methodology}
\input{results}
\input{end}

\bibliographystyle{IEEEtran}
\bibliography{citation}
% ─────────────────────────────────────────────────────────────

\end{document}
