---
title: "Insurance"
author: "Doan Cong Thanh"
date: "2024-07-31"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE,warning = FALSE, message = FALSE)
```

## Load Libraries and Data
```{r}
# Load necessary libraries
library(car)
library(caret)
library(dplyr)
library(MASS)
# Load the data set
data <- read.csv("insurance.csv")
```
## Encode Categorical Variables
```{r}
# Encode the data
# Create pr_sex variable: 0 for female, 1 for male
data$pr_sex <- ifelse(data$sex == "male", 1, 0)

# Create pr_smoker variable: 0 for non-smoker, 1 for smoker
data$pr_smoker <- ifelse(data$smoker == "yes", 1, 0)

# Create dummy variables for region, excluding 'northwest' as the reference group
data$pr_region_northeast <- ifelse(data$region == "northeast", 1, 0)
data$pr_region_southeast <- ifelse(data$region == "southeast", 1, 0)
data$pr_region_southwest <- ifelse(data$region == "southwest", 1, 0)
```
## Handle data
```{r}
# Fit the initial model
initial_model <- lm(charges ~ pr_sex + pr_smoker + pr_region_northeast + pr_region_southeast + pr_region_southwest + age + bmi + children, data = data)

# Check for normality of residuals using Shapiro-Wilk test
shapiro_test <- shapiro.test(residuals(initial_model))
print(shapiro_test)

# Draw boxplots for all numeric columns to visualize outliers
numeric_columns <- sapply(data, is.numeric)
par(mfrow=c(2,4))
for(i in which(numeric_columns)){
  boxplot(data[,i], main=names(data)[i])
}

# Calculate Cook's distance
cooksd <- cooks.distance(initial_model)

# Identify influential points (top 10% by Cook's distance)
influential <- as.numeric(names(sort(cooksd, decreasing = TRUE)[1:floor(0.1 * length(cooksd))]))

# Remove influential observations
data_clean <- data[-influential, ]

# Check for NA values and remove them
if(any(is.na(data_clean))){
  data_clean <- na.omit(data_clean)
}

# Draw boxplots for all numeric columns to visualize outliers after cleaning
numeric_columns <- sapply(data_clean, is.numeric)
par(mfrow=c(2,4))
for(i in which(numeric_columns)){
  boxplot(data_clean[,i], main=names(data_clean)[i])
}

# Percentage of data removed
(dim(data)[1] - dim(data_clean)[1]) / dim(data)[1] * 100

# Fit a linear model on the cleaned data to check for normality again
clean_model <- lm(charges ~ pr_sex + pr_smoker + pr_region_northeast + pr_region_southeast + pr_region_southwest + age + bmi + children, data = data_clean)
shapiro_test_cleaned <- shapiro.test(residuals(clean_model))
print(shapiro_test_cleaned)
```
## Calculate VIF
```{r}
# Calculate VIF
vif_data <- vif(lm(charges ~ pr_sex + pr_smoker + pr_region_northeast + pr_region_southeast + pr_region_southwest + age + bmi + children, data = data_clean))
print(vif_data)
```
The VIF table indicates that the VIF values of six predictors are nearly close to 1, which means there is no severse multicollinearity problem among this model.

## Split Data into Training and Testing Sets
```{r}
# Set seed for reproducibility
set.seed(123)

# Split data
train_index <- createDataPartition(data_clean$charges, p = 0.8, list = FALSE)
train_data <- data_clean[train_index, ]
test_data <- data_clean[-train_index, ]
```

## Train Linear Regression Model
### Define Full and Initial Models
```{r}
# Define the full model
modFull <- lm(charges ~ pr_sex + pr_smoker + pr_region_northeast + pr_region_southeast + pr_region_southwest + age + bmi + children, data = train_data)

# Define the null model
modZero <- lm(charges ~ 1, data = train_data)

# Define an intermediate model (you can customize the variables as needed)
modInter <- lm(charges ~ pr_sex + pr_smoker + pr_region_northeast + age + bmi, data = train_data)
```
### Apply Step-Wise Method
```{r}
# Backward elimination with AIC
modBest1A <- stepAIC(modFull, direction = "backward", scope = list(lower = modZero, upper = modFull), k = 2)

# Backward elimination with BIC
modBest1B <- stepAIC(modFull, direction = "backward", scope = list(lower = modZero, upper = modFull), k = log(nrow(data)))

# Forward selection with BIC
modBest2 <- stepAIC(modZero, direction = "forward", scope = list(lower = modZero, upper = modFull), k = log(nrow(data)))

# Stepwise selection with BIC
modBest3 <- stepAIC(modInter, direction = "both", scope = list(lower = modZero, upper = modFull), k = log(nrow(data)))
```
These 4 models are exactly one. So we choose the modBest1A.
### Summary of Best Model
```{r}
summary(modBest1A)
```
## Evaluate Model Performance
```{r}
# Predict on test data
predictions <- predict(modBest1A, newdata = test_data)

# Calculate performance metrics
actual <- test_data$charges
mse <- mean((predictions - actual)^2)
rmse <- sqrt(mse)
r_squared <- cor(predictions, actual)^2

# MSE
mse
# RMSE
rmse
# R-squared
r_squared
```

### Mean Squared Error (MSE): 14,044,199

Interpretation: MSE is the average of the squares of the errors between predicted and actual values. It gives a rough idea of how large the errors are. In this case, the MSE is quite large, indicating that there are significant differences between the predicted and actual values.

### Root Mean Squared Error (RMSE): 3,747.559

Interpretation: RMSE is the square root of MSE and provides a measure of the average magnitude of the prediction errors in the same units as the response variable. An RMSE of 3,747.559 means that the typical error in predicting the charges is about 3,747.559 units.

### R-squared (R²): 0.9160806

Interpretation: R-squared represents the proportion of the variance in the dependent variable that is predictable from the independent variables. An R-squared value of 0.9160806 means that approximately 91.61% of the variability in the charges can be explained by the model. This is a high value, indicating that the model fits the data well.

## Comments on Model Performance

### High R-squared
The R-squared value of 0.9160806 suggests that the model explains a large portion of the variance in the response variable (charges). This implies that the predictors used in the model are quite effective in capturing the relationship between the inputs and the output.

### High RMSE
While the R-squared value is high, the RMSE value of 3,747.559 indicates that there is still a considerable amount of error in the predictions. This suggests that while the model captures the overall trend well, individual predictions may still be off by a significant margin.

### Model Appropriateness
Given the high R-squared value, the model seems to be a good fit for the data. However, the high RMSE indicates that there might be room for improvement, possibly by refining the model, adding more predictors, or considering non-linear relationships.

# Apply Box-Cox transformation
```{r}
# Fit a linear model on the training data
initial_model <- lm(charges ~ pr_sex + pr_smoker + pr_region_northeast + pr_region_southeast + pr_region_southwest + age + bmi + children, data = train_data)

# Find the optimal lambda for Box-Cox transformation
boxcox_result <- boxcox(initial_model, lambda = seq(-2, 2, by = 0.1))
lambda_optimal <- boxcox_result$x[which.max(boxcox_result$y)]

# Apply the Box-Cox transformation
if(lambda_optimal == 0) {
  train_data$charges_transformed <- log(train_data$charges)
  test_data$charges_transformed <- log(test_data$charges)
} else {
  train_data$charges_transformed <- (train_data$charges^lambda_optimal - 1) / lambda_optimal
  test_data$charges_transformed <- (test_data$charges^lambda_optimal - 1) / lambda_optimal
}

# Fit the model with the transformed response variable
modBestTransformed <- lm(charges_transformed ~ pr_sex + pr_smoker + pr_region_northeast + pr_region_southeast + pr_region_southwest + age + bmi + children, data = train_data)

# Summary of the transformed model
summary(modBestTransformed)

# Evaluate model performance on test data with transformed response variable
predictions_transformed <- predict(modBestTransformed, newdata = test_data)
actual_transformed <- test_data$charges_transformed

# Calculate performance metrics for transformed model
mse_transformed <- mean((predictions_transformed - actual_transformed)^2)
rmse_transformed <- sqrt(mse_transformed)
r_squared_transformed <- cor(predictions_transformed, actual_transformed)^2

# MSE for transformed model
mse_transformed
# RMSE for transformed model
rmse_transformed
# R-squared for transformed model
r_squared_transformed
```

